# Analysis of numeric data

```{r include=F}
library(tidyverse)
library(magrittr)
```

So far, we have dealt with descriptive statistics and the analysis of the sample 
data collected. However, the bane of most statistical analysis is to make 
inferences about the population as a whole. In this section, we mainly do an 
inferential analysis of continuous variables.

## Confidence interval of a mean
To determine the confidence interval of the mean of a numeric variable in R, we 
use the One Sample Student's T-test. The assumptions for the validity of this 
test are:

1. The sample should have been randomly chosen
1. The population distribution of the variable should be normal. This can be 
assumed to be present if
    - The distribution of the population is known to be normally distributed
    - The population distribution should have one mode, symmetric, without 
    outliers and a sample size of 15 or less
    - The population distribution should be moderately skewed, without outliers, 
    and have one mode and with a sample size between 16 and 40
    - The sample size is more than 40 and the data has no outliers.

With our sample considered to be randomly selected and a sample size of 140, we 
apply the One-sample T-test as below. 

We first import the data 
```{r}
df_data1 <- 
    read_delim(
        file = "./Data/data1.txt", 
        delim = "\t",
        col_types = c("c", "f", "i","i")
    )

glimpse(df_data1)
```



```{r}
df_data1 %$% 
    epiDisplay::ci(height)%>% 
    knitr::kable(booktabs = TRUE)
```

```{r}
df_data1 %>% 
    meantables::mean_table(height) %>% 
    knitr::kable(booktabs = TRUE)
```

For sex stratified confidence intervals we have

```{r}
df_data1 %>% 
    group_by(sex) %>% 
    meantables::mean_table(height)%>% 
    knitr::kable(booktabs = TRUE)
```

## Comparing the mean to a hypothesised value
Assuming the objective of the data collected  was to determine if the average 
weight of our population was similar to a population with a known mean weight of 
14kgs. 

Our null hypothesis is: 

> H0: There is no difference in mean weight between our population and a 
population with mean weight of 14kgs.

To test this hypothesis we use the One sample t-test after we have satisfied 
ourselves that the assumptions for its use have been met.

```{r}
df_data1 %>% 
     rstatix::t_test(formula = weight~1, mu=14) %>% 
     knitr::kable(booktabs = TRUE)
```

The p-value of 0.001 is the probability that our sample could have come from a 
population with a mean weight of 14kgs. Since this is very small we reject the 
null H0 at a 5% significance level and conclude that our population mean weight 
is significantly different from 14kgs.
The confidence interval generated is that of our sample mean. With the 
hypothesized value of
14kgs outside the confidence interval of the mean we conclude that there is 
insuficient evidence to suggest that the mean weight of our population is 14kgs.

## Comparing mean of two independent groups
This is possibly the most common use for the t-test. To compare the mean weights 
of the males and female in our study we come up with

> H0: There is no difference in weight between the males and females in our 
population.

To test this assertion we first determine if our sample fits the assumption for 
the use of the Two sample t-test. These are:

1. The sample should have been randomly chosen
1. The two samples are completely independent
1. Each population is at least 20 times larger than its respective sample.
1. The population distribution of the variable should be normal. This can be assumed to be
present if
    - The distribution of the population is known to be normal
    - The population distribution should have one mode, symmetric, without outliers and
a sample size of 15 or less
    - The population distribution should be moderately skewed, without outliers, have
one mode and with a sample size between 16 and 40
    - The sample size is more than 40 and data has no outliers.
    
Our data fulfils all the criteria above hence we apply the test

```{r}
df_data1 %>% 
    rstatix::t_test(formula = weight ~ sex, detailed = T)%>% 
    knitr::kable(booktabs = TRUE)
```

With a relatively high p-value we conclude that there is insufficient evidence 
to refute null hypothesis. In other words there is insufficient evidence to 
conclude the mean weights of males and females differ in the study population. 
Note that in our sample however females appear heavier than males as shown by 
the last two lines of the output above.

The confidence interval determined above (-0.86 to 3.47) is actually that for 
the mean sample difference between females and males. Since the confidence 
interval contains the null value from H0 above i.e. 0, we conclude that there 
isn't enough evidence of a difference in mean weight between the two sexes.
Hence both the confidence interval and p-value come to similar conclusions.

## Comparing means of paired observations
In this section we use the `bread.txt` data which has the weight in grams before 
and after baking some loaves of bread. A description of the variables is 
contained in the data file. Paired observations occur in circumstances where 
repeated measurement are done on the same object or data collected has some 
characteristics in common. For the bread data the same bread is weighed before 
and after baking. Determining if there is a significant difference between the 
two measurements requires the use of the Paired t-test. As always be begin by 
importing the data

Next we determine the structure of the data frame df2

```{r}
df_bread <-read.table("./Data/bread.txt", sep="\t", header=T)
```

And then summarize it 

```{r}
df_bread %>% summarytools::dfSummary(graph.col = F)
```

Our next task is to compare the weight of the loaves of bread before and after 
baking. We begin by looking at the mean and standard deviations of the two 
weights.

```{r}
df_bread %>% 
    select(before, after) %>% 
    rstatix::get_summary_stats(type= "mean_sd")
```

It is obvious that the mean weight of the bread before baking is much higher 
than after however the standard deviations appear similar. For a formal test to 
determine the difference in means we use the paired t-test. As before the we 
state the assumptions of the paired t-test 

1. The sample should have been randomly chosen
1. The two samples are not independent (they are related)
1. Each population is at least 20 times larger than its respective sample.
1. The population distribution of the difference between the two variables 
should be normal.
This can be assumed to be present if
    - The distribution of the population is known to be normal
    - The population distribution should have one mode, symmetric, without 
    outliers and a sample size of 15 or less
    - The population distribution should be moderately skewed, without outliers, 
    have one mode and with a sample size between 16 and 40
    - The sample size is more than 40 and data has no outliers.
    
The only new assumption we need to evaluate here is the distribution of the 
difference between the weights before and after. We do so below

```{r fig.height=3, fig.width=5}
df_bread %>% 
    mutate(diff_in_wgt = after - before) %>% 
    ggplot(aes(x = diff_in_wgt)) +
    geom_histogram(bins = 10, col = "white") + 
    labs(x = "Difference in weight") +
    theme_bw()
```

Alternatively, we can perform a Shapiro-Wilk's test for normality. This has H0 
as not deviating from the normal distribution. This is done below
```{r}
df_bread %>% 
    mutate(diff_in_wgt = after - before) %>%
    rstatix::shapiro_test(vars = "diff_in_wgt")
```


The output and graphical representation above shows our difference in weight is 
literally normally distributed. We therefore go ahead to determine the 
difference in mean weights. First we state our hypothesis

> H0: There is no change in weight of loaves of bread after baking

And then perform the test after converting the data to the long format

```{r}
df_bread %>%
    pivot_longer(
        cols = c(before, after), 
        names_to = "time", 
        values_to = "weight"
    ) %>% 
    rstatix::t_test(
        formula = weight~time, 
        paired = TRUE, 
        detailed = TRUE
    )%>% 
    knitr::kable(booktabs = TRUE)
```

There was on average a 162.5g reduction in weight of the loaves of bread after 
baking. This reduction has a 95% confidence interval of 159.1g to 165.9g and is 
significantly different from 0 (p-value<0.001).

## Test for equality of variances
In using the Student's T-test to determine difference between the means of two 
independent groups we need to be mindful of the variances of each group. The 
computations done for the independent groups t-test are different when the 
variances between the groups are similar or different. Therefore to determine 
if the mean weight significantly differ between males and females we need to 
determine and compare their variances. The function `var.test()` in R compares 
the variances between two independent groups and can be used for this 
determination.Below we apply this F-test to compare the variances of the weight 
for the two sexes. First we determine the variances.

```{r}
df_data1 %>% 
    group_by(sex) %>%
    summarise(across(weight, list(var = var, meam = mean)))
```

There seem to be a big difference between the variance of the weights for the 
two sexes. That for the females is almost 1.7 times that of the males. To 
determine if this is not a chance finding we apply a formal statistical test. 
Here our

> H0: There is no difference in the variance of the weights for males and 
females in our population

The F-test actually tests the ratio of the variances not the difference. In that 
regard our null value would be 1.

```{r}
df_data1 %>% 
    var.test(formula = weight~sex, data = .) %>% 
    broom::tidy() %>% 
    knitr::kable(booktabs = TRUE)
```

The significant p-value (at a significance level of 0.05) and a confidence 
interval not containing 1 (the null value) implies there is very little evidence 
that the variance between the two groups are the same (in other words they 
differ significantly). In that case the conclusion from the previous analysis is 
valid as R assumes the variances to be unequal by default if the `t.test()`
function is used.

Next we apply the same principle to the determine if the mean heights are 
similar for males and females in our population. We first determine if the 
variances are significantly different.

```{r}
df_data1 %>% 
    group_by(sex) %>%
    summarise(across(height,
                     list(
                         var = ~var(., na.rm=T),
                         meam = ~mean(., na.rm=T))))
```

From the results above the variance for the females look much higher (1.3 times) 
than the males however we apply a test to formally determine this. 

```{r}
df_data1 %>% 
    var.test(formula = height~sex, data = .) %>% 
    broom::tidy() %>% 
    knitr::kable(booktabs = TRUE)
```

Both p-value and confidence interval conclude there is insufficient evidence to 
say the two variances are different. To use the `t.test()` function to determine 
the possibility that mean height differ between males and females we specify 
that variance is equal as below.

```{r}
df_data1 %>%
    rstatix::t_test(
        formula = height~sex, 
        var.equal = TRUE, 
        detailed = TRUE
    )%>% 
    knitr::kable(booktabs = TRUE)
```



