# Analysis of numeric data

```{r include=F}
library(tidyverse)
library(magrittr)
```

So far, we have dealt with descriptive statistics and the analysis of the sample 
data collected. However, the bane of most statistical analysis is to make 
inferences about the population as a whole. In this section, we mainly do an 
inferential analysis of continuous variables.

## Confidence interval of a mean
To determine the confidence interval of the mean of a numeric variable in R, we 
use the One Sample Student's T-test. The assumptions for the validity of this 
test are:

1. The sample should have been randomly chosen
1. The population distribution of the variable should be normal. This can be 
assumed to be present if
    - The distribution of the population is known to be normally distributed
    - The population distribution should have one mode, symmetric, without 
    outliers and a sample size of 15 or less
    - The population distribution should be moderately skewed, without outliers, 
    and have one mode and with a sample size between 16 and 40
    - The sample size is more than 40 and the data has no outliers.

With our sample considered to be randomly selected and a sample size of 140, we 
apply the One-sample T-test as below. 

We first import the data 
```{r}
df_data1 <- 
    read_delim(
        file = "./Data/data1.txt", 
        delim = "\t",
        col_types = c("c", "f", "i","i")
    )

glimpse(df_data1)
```



```{r}
df_data1 %$% 
    epiDisplay::ci(height)%>% 
    flextable::flextable()
```

```{r}
df_data1 %>% 
    meantables::mean_table(height) %>% 
    flextable::flextable()
```

For sex stratified confidence intervals we have

```{r}
df_data1 %>% 
    group_by(sex) %>% 
    meantables::mean_table(height)%>% 
    flextable::flextable()
```


To test the H0: The mean height is 100cm whiles generating the confidence 
interval as well we use