
```{r include =F}
library(huxtable)
library(gtsummary)
library(tidyverse)
library(magrittr)
```

# Confounding and Interaction

## Introduction
A new waiter employed in a bar to serve ice in glasses in Accra made an 
observation which seem irrational. He realized that customers who he served 
ice often ended up drunk. He is the first to accept this does not make sense but 
strongly believe he is right. He approached a friend who is a researcher and 
they both decided to investigate this. The friend decided to take data on 510
randomly selected customers. This data is recorded in `drinks.txt`.

The variables collected include id the sequentially allocated study id, sex, 
sex of the `customer`, `liquor` whether the customer took an alcoholic liquor 
(spirits, brandy, etc), `ice` whether the customer was served ice in a glass, 
`drunk` whether the customer ended up being drunk, `food` whether the customer 
was served food and `age`, age in years of the customer. The task now is to 
determine from the data if really ingesting ice is associated with being drunk.

## Effect and effect size
### Effect
Before we begin to discuss the above problem we first look at what in 
statistical terms is meant by effect. Literally effect is defined as a change 
that occurs as a consequence of an action. In statistical terms effect is 
usually the change in one variable by another. For instance, effect could be the 
mean change in blood pressure (effect) after taking a drug (the action). Two of
the most common ways of expressing effect in categorical data analysis are the 
odds ratio (OR) and Relative Risk (RR). These two usually determine effect by 
comparing the odds or risk in two groups and expressing the effect as the 
ratios.

### Effect size
It does not only suffice to report the effect but also its size. People would 
want to know not only if an intervention makes a difference but also how much 
difference it makes. Effect size can be in opposite direction. For instance the 
difference between the blood pressure after taking a medication may be positive 
(when the pressure after taking the drug is more than before taking) or negative 
(when the pressure becomes less when drug is taken). On the other hand effects 
can be on the same side but with a difference in magnitude. On OR of 5.1 is much 
higher than 1.5 despite both indicating an increased odds of having the outcome. 
We are going to apply this notion in determining confounders and effect 
modifiers.

Now back to the waiter's issue. First we read in the data.

```{r}
df_drinks <- 
    read.table("./Data/drinks.txt", header=T, sep="\t") %>% 
    mutate(
        across(
            c(liquor, ice, drunk, food), 
            ~factor(.x, levels = c(0,1), labels = c("No","Yes"))
        ),
        sex = factor(sex)
    )
```

And then summarize the data

```{r}
options(huxtable.knit_print_df = FALSE)
df_drinks %>% 
    summarytools::dfSummary(graph.col = F)

options(huxtable.knit_print_df = TRUE)
```
In this section we make use of two main functions cc() and mhor() from the 
`epicalc` package. First we determine the effect of having ice with respect to 
being drunk by using the odds ratio.

```{r, fig.height=5, fig.width=6}
df_drinks %$% epiDisplay::cc(drunk, ice)
```

It appears the waiter is right! Taking ice is associated with a significantly 
higher odds of getting drunk (OR: 1.73, 95% CI: 1.17 to 2.55, p<.001). This we 
all agree defies logic. The question then is: How did this arise?

### Confounding
In research there often arises the situation where the observed effect of a 
variable tend to depend on another. For instance in our case the effect of ice 
in making customers drunk could be because it is associated with another item. 
If customers who took ice also took in some hard liquor at the same time this 
may make it seem that taking ice is associated with being drunk. This would be 
a typical case of confounding with the liquor ingestion 
confounding the effect of ice. 

For a variable to be a confounder it must meet these basic properties

1. The confounder must be related to the exposure (ice) variable
1. The confounder must be related to the outcome (drunk) variable
3. The confounder should not be on the causal pathway between the exposure(ice) 
and outcome (drunk).

The first two conditions are easily tested statistically but the last can only 
be tested with prior knowledge, often coming from epidemiological and scientific 
facts. The investigator of the bar study suspects liquor to be a confounder to 
the effect of ice so sets out to determine this. First he looks out for the 
relationship between the suspected confounder and the outcome

```{r fig.height=5, fig.width=6}
df_drinks %$% epiDisplay::cc(drunk, liquor)
```

Obviously having liquor is strongly associated with being drunk with an 
OR = 2.41 (95%CI: 1.6, 3.62, p<.001). Next we test the association between the 
exposure and the possible confounder.

```{r fig.height=5, fig.width=6}
df_drinks %$% epiDisplay::cc(ice, liquor, graph = F)
```

There is a strong association between having ice and taking hard liquor 
(OR: 13.71, 95% CI: 8.48 to 22.33, p<.001). With this relationship established 
the first two conditions stated above have been fulfilled.

Our next task is to determine if consumption of liquor is a confounder to the 
effect of ice. We do this by determining the adjusted OR and comparing it to the 
crude OR (unadjusted OR). We recall from above that the crude OR is 1.73 
(95% CI = 1.17 to 2.55, p=0.004). The Mantel-Haenszel odd ratios determined by 
the mhor() function is an OR adjusted for the possible confounder. This is shown 
below

```{r fig.height=5, fig.width=6}
df_drinks %$% epiDisplay::mhor(drunk, ice, liquor)
```

The function first stratifies the odds of getting drunk after taking ice into 
the individual groups of the possible confounding variable. It then determines 
the OR for each level of the confounder, 1.14(95%CI= 0.621 to 2.05, p=0.666) for 
those who took liquor and 1.13 (95%CI: 0.497 to 2.58, p=0.848) for those who did 
not. It then reports the MH combined (adjusted) OR of 1.14 (95%CI: 0.726 to 
1.78, p=0.574). Thus stratification or adjustment has reduced the OR or effect of
ice from a significant 1.7 to a non-significant relationship 1.1. The effect has 
all but disappeared as ice is no longer significantly associated with getting 
drunk after adjustment. This shows that ice on its own does not cause people to 
be drunk and the observed effect was just an issue of confounding by the 
consumption of liquor.

Above we have shown the relationship between the three variables involved in 
demonstrating confounding effect of liquor on the effect of taking ice. We can 
then ask: Is ice a confounder to the relationship between being drunk and 
consumption of liquor? Remember the crude OR for this effect is 2.41 (95%CI: 
1.6, 3.62, p<.001). To answer the question we determine the adjusted OR below.

```{r fig.height=4, fig.width=6}
df_drinks %$% epiDisplay::mhor(drunk, liquor, ice)
```

The output shows that the adjusted OR (OR=2.24, 95%CI: 1.41 to 3.56, p<.001) is 
barely different from the crude OR and also retains its significant effect. The 
consumption of ice is therefore not a confounder to the effect of hard liquor.

In summary to determine if a variable is a confounder we need to determine its 
association with the exposure and outcome variable. If there is an association 
we then determine the crude effect and adjusted effect. If these two are 
significantly different then we can conclude the presence of a confounder.

However, the definition of "significant effect" is always difficult to determine. 
Some say a change in at least 10% of the crude effect can be considered enough. 
Whenever there is a possible confounder the adjusted effect should be reported. 
Also, it is worth noting that effect change could be making the effect higher or 
lower. A confounder that for instance results in a crude OR of 2.4 and an 
adjusted OR of 1.5 is a positive confounder while one that has a crude OR of
1.5 but adjusted OR of 2.5 shows negative confounding.

## Interaction or Effect modification
It is common knowledge among those who drink alcoholic beverages that filling 
your tummy with food often delays getting drunk when one consumes alcoholic 
beverages. The investigator of the bar data now wants to determine if this is 
actually so. If this notion is right then getting drunk after taking hard liquor 
should be dependent on whether the customer also ordered and ate some food as 
well. This is a typical example of effect modification (a term often used by 
epidemiologist) or interaction (used by statisticians). The two essentially 
means the same. By definition effect modification occurs when the effect size of 
an exposure (liquor) on an outcome (drunk) differs depending on the level of a 
third variable (eating food). When this occurs just computing and reporting the 
overall effect is misleading.

We begin our investigation for food causing an interaction in the effect of 
drinking hard liquor by determining the OR of getting drunk when one took some 
food and when one didn't separately.

We begin with those who ate some food.

```{r}
df_drinks %>% 
    filter(food == "Yes") %$% 
    epiDisplay::cc(drunk, liquor)
```

There is a statistically insignificant odds of getting drunk after taking hard 
liquor when one eats as well with an OR 1.41 (95% CI: 0.8 to 2.51, p=0.208). 
Next we do same for those who did not eat

```{r fig.height=4, fig.width=6}
df_drinks %>% 
    filter(food == "No") %$% 
    epiDisplay::cc(drunk, liquor)
```


The OR of getting drunk if one takes liquor and on an empty stomach is much 
higher and statistically significant (OR: 3.38, 95% CI: 1.82 to 6.26, p<.001). 
We seem to have quite a substantial difference in effect of drinking liquor 
depending on food intake or not. To be sure if this food intake is a significant 
effect modifier we subject this to a formal test using the `mhor()` function.

```{r fig.height=4, fig.width=6}
df_drinks %$% epiDisplay::mhor(drunk, liquor, food, graph = F)
```

The stratified analysis output above shows our prior calculated stratified ORs 
as well as the combined Mantel-Haenszel OR. The Homogeneity test above tests the 
null hypothesis:

> H0: There is no difference between the two stratified Odds ratios.

A p-value of 0.031 therefore indicates a statistically significant difference 
in effects (OR) for those who ate and those who did not. We therefore conclude 
that there is significant interaction between the food intake and getting drunk 
if one takes hard liquor.

In this section we have dealt with detecting a confounder or interaction variable 
in categorical data analysis. This is just scratching the surface. In subsequent 
chapters we would be dealing a lot more with these but using regression analysis.
