```{r echo=FALSE, message=F}
library(tidyverse)
library(magrittr)
library(gtsummary)
library(huxtable)
mytheme <- function(ht){
    ht <- huxtable::theme_article(ht)
    ht <- huxtable::set_top_padding(ht, row = "2pt")
    ht <- huxtable::set_bottom_padding(ht, row = "2pt")
    ht <- huxtable::set_font(ht, value = "Arial, Times New Roman, Times, Serif")
    ht <- huxtable::set_font_size(ht, value = 11)
}
options(huxtable.knit_print_df_theme = mytheme)
```

# Descriptive Statistics: Numeric

For this section, we will use the `NewDrug_clean.dta` dataset

```{r comment=NA}
newdrug <-  
    haven::read_dta("./Data/NewDrug_clean.dta") %>% 
    mutate(sex  = haven::as_factor(sex), treat = haven::as_factor(treat)) %>% 
    haven::zap_labels() 

newdrug %>% summary()
```


### Single continuous variable

#### Measures of Central Tendency & Dispersion

These include mean and median and are displayed below

```{r comment=NA}
newdrug %>% 
    summarise(
        mean.bp1 = mean(bp1), 
        median.bp1 = median(bp1), 
        sd.bp1 = sd(bp1), 
        min.bp1 = min(bp1), 
        max.bp1 = max(bp1),
        iqr = IQR(bp1)
    ) 

```

Alternatively

```{r comment=NA}
newdrug %$% 
    psych::describe(bp1)
```

And to show the interquartile range we do the following.

```{r comment=NA}
newdrug %$% 
    psych::describe(bp1, IQR = TRUE,quant = c(.25, .75))
```

#### Graphs - Histogram 
```{r comment=NA, fig.width = 6, fig.height=4}
newdrug %>% 
    ggplot(aes(x = bp1)) + 
    geom_histogram(bins = 7, col="black", alpha = .5, fill = "red") +
    labs(title = "Histogram of Blood Pressure before  intervention",
         x= "BP1")+
    theme_light()
```

#### Graphs - Boxplot and violin plot 

```{r comment=NA, fig.width=6, fig.height=4}
newdrug %>% 
    ggplot(aes(y = bp1)) + 
    geom_boxplot(col="black",  
                 alpha = .2, 
                 fill = "blue", 
                 outlier.fill = "black",
                 outlier.shape = 22) +
    labs(title = "Boxplot of Blood Pressure before  intervention",
         y = "BP1")+
    theme_light()
```

#### Graphs - Density plot
```{r comment=NA, fig.width = 6, fig.height=4}
newdrug %>% 
    ggplot(aes(y = bp1)) + 
    geom_density(col="black", fill = "yellow", alpha=.6) +
    labs(title = "Density Plot of Blood Pressure before  intervention",
         y = "Blood Pressure before  intervention")+
    coord_flip() +
    theme_light()
```

#### Graphs - Cumulative Frequency plot

```{r comment=NA, fig.width = 6, fig.height=4}
newdrug %>% 
    group_by(bp1) %>% 
    summarize(n = n()) %>% 
    ungroup() %>% 
    mutate(cum = cumsum(n)/sum(n)*100) %>% 
    ggplot(aes(y = cum, x = bp1)) +
    geom_line(col=3, linewidth=1.2)+
    labs(
        title = "Cumulative Frequency Plot of Blood Pressure before  intervention",
        x = "BP1",
        y = "Cumulative Frequency")+
    theme_light() 

```

### Multiple Continuous variables

#### Measures of Central tendency & Dispersion

```{r comment=NA}
newdrug %>% 
    select(where(is.numeric)) %>% 
    psych::describe()
```

To illustrate graphing multiple continuous variables we use the 2 bp variables

```{r comment=NA}
bps <- 
    newdrug %>%
    select(bp1, bp2) %>% 
    pivot_longer(
        cols = c(bp1, bp2),
        names_to = "measure", 
        values_to = "bp") %>% 
    mutate(
        measure = fct_recode(
            measure, "Pre-Treatment" = "bp1", "Post-Treatment" = "bp2"
            )
        )
```

Next, we create multiple density plots

```{r comment=NA, fig.width = 6, fig.height=4}
bps %>% 
    ggplot(aes(y = measure, x = bp, fill = measure)) +
    ggridges::geom_density_ridges2( col="black", alpha = .5, scale=1, 
                                    show.legend = F) +
    labs(x = "Blood pressure (mmHg)", 
         y = "Density", 
         fill = "Blood Pressure") +
    theme_bw()
```


```{r comment=NA, fig.width = 6, fig.height=4}
bps %>% 
    ggplot(aes(y = measure, x = bp, fill = measure))+
    geom_boxplot(show.legend = FALSE) +
    labs(y = NULL, 
         x = "Blood Pressure", 
         fill = "Blood Pressure") +
    coord_flip()+
    theme_light() 
```


```{r comment=NA, fig.width = 6, fig.height=4}
bps %>% 
    ggplot(aes(y = measure, x = bp, fill = measure))+
    geom_violin(show.legend = FALSE) +
    coord_flip()+
    theme_light() 
    
```

## Continuous by a single categorical variable
### Summary

We do this with one variable.
```{r comment=NA}
newdrug %>% 
    group_by(treat) %>% 
    summarize(mean.bp1 = mean(bp1),
              sd.bp1 = sd(bp1),
              var.bp1 = var(bp1),
              se.mean.bp1 = sd(bp1)/sqrt(n()),
              median.bp1 = median(bp1),
              min.bp1 = min(bp1),
              max.bp1 = max(bp1)) %>% 
    ungroup()
```


### Graph - Histogram, Boxplot, Density plot and cumulative frequency
The graphs are similar to the above so we skip them. 

## Continuous by multiple categorical variables
### Summary
This can be done as below.

```{r comment=NA}
newdrug %>% 
    group_by(treat, sex) %>% 
    summarize(mean.bp1 = mean(bp1),
              sd.bp1 = sd(bp1),
              var.bp1 = var(bp1),
              se.mean.bp1 = sd(bp1)/sqrt(n()),
              median.bp1 = median(bp1),
              min.bp1 = min(bp1),
              max.bp1 = max(bp1),
              .groups = "drop") 
```

And this can be presented in a boxplot below

```{r fig.width=6, fig.height=4}
newdrug %>% 
    ggplot(aes(y = bp1, x = sex, fill = treat)) +
    geom_boxplot()+
    labs(
        y = "Blood Pressure (mmHg)",
        x =  "Sex",
        fill = 'Treatment') +
    theme_bw()
```

