```{r echo=FALSE, message=F}
library(tidyverse)
library(magrittr)
library(gtsummary)
library(huxtable)
mytheme <- function(ht){
    ht <- huxtable::theme_article(ht)
    ht <- huxtable::set_top_padding(ht, row = "2pt")
    ht <- huxtable::set_bottom_padding(ht, row = "2pt")
    ht <- huxtable::set_font(ht, value = "Arial, Times New Roman, Times, Serif")
    ht <- huxtable::set_font_size(ht, value = 11)
}
options(huxtable.knit_print_df_theme = mytheme)
```

# Descriptive Statistics: Numeric

The initial analysis of numeric data is usually a description of the data at hand without making inference to the population from which the data was drawn. This gives the data analyst a general overview of the data at hand, how best to describe it and what analysis best suits it. In descriptive analysis of numeric data the most basic is to determine the:

- Measure of Central Tendency: This is a description of the center of the data. These measures include mean, median and mode.
- Measure of Dispersion: A measure of how widespread the data is. These include standard deviation, variance, interquartile range and range.

For this section, we will use the `NewDrug_clean.dta` dataset

```{r comment=NA}
newdrug <-  
    haven::read_dta("./Data/NewDrug_clean.dta") %>% 
    mutate(sex  = haven::as_factor(sex), treat = haven::as_factor(treat)) %>% 
    haven::zap_labels() 

newdrug %>% summary()
```

## Single continuous variable

### Measures of Central Tendency & Dispersion

Below we determine the mean, median, standard deviation, range (minimum, maximum) and interquartile range of out initial blood pressure

```{r comment=NA}
newdrug %>% 
    summarise(
        Mean = mean(bp1), 
        Median = median(bp1), 
        Standard_Dev = sd(bp1), 
        Minimum = min(bp1), 
        Maximum = max(bp1),
        IQR = IQR(bp1)
    ) 
```

Alternatively, the `psych` package gives these measures in further details. The output includes a measure of the Kurtosis and Skewness, both describing the shape of the data.

```{r comment=NA}
newdrug %$% 
    psych::describe(bp1)
```

And to show the interquartile range we do the following.

```{r comment=NA}
newdrug %$% 
    psych::describe(bp1, IQR = TRUE,quant = c(.25, .75))
```

### Graphs - Histogram 
```{r comment=NA, fig.width = 6, fig.height=4}
newdrug %>% 
    ggplot(aes(x = bp1)) + 
    geom_histogram(bins = 7, col="black", alpha = .5, fill = "red") +
    labs(title = "Histogram of Blood Pressure before  intervention",
         x= "BP1")+
    theme_light()
```

### Graphs - Boxplot and violin plot 

```{r comment=NA, fig.width=6, fig.height=4}
newdrug %>% 
    ggplot(aes(y = bp1)) + 
    geom_boxplot(col="black",  
                 alpha = .2, 
                 fill = "blue", 
                 outlier.fill = "black",
                 outlier.shape = 22) +
    labs(title = "Boxplot of Blood Pressure before  intervention",
         y = "BP1")+
    theme_light()
```

#### Graphs - Density plot
```{r comment=NA, fig.width = 6, fig.height=4}
newdrug %>% 
    ggplot(aes(y = bp1)) + 
    geom_density(col="black", fill = "yellow", alpha=.6) +
    labs(title = "Density Plot of Blood Pressure before  intervention",
         y = "Blood Pressure before  intervention")+
    coord_flip() +
    theme_light()
```

#### Graphs - Cumulative Frequency plot

```{r comment=NA, fig.width = 6, fig.height=4}
newdrug %>% 
    group_by(bp1) %>% 
    summarize(n = n()) %>% 
    ungroup() %>% 
    mutate(cum = cumsum(n)/sum(n)*100) %>% 
    ggplot(aes(y = cum, x = bp1)) +
    geom_line(col=3, linewidth=1.2)+
    labs(
        title = "Cumulative Frequency Plot of Blood Pressure before  intervention",
        x = "BP1",
        y = "Cumulative Frequency")+
    theme_light() 

```

### Multiple Continuous variables

#### Measures of Central tendency & Dispersion

```{r comment=NA}
newdrug %>% 
    select(where(is.numeric)) %>% 
    psych::describe()
```

To illustrate graphing multiple continuous variables we use the 2 bp variables

```{r comment=NA}
bps <- 
    newdrug %>%
    select(bp1, bp2) %>% 
    pivot_longer(
        cols = c(bp1, bp2),
        names_to = "measure", 
        values_to = "bp") %>% 
    mutate(
        measure = fct_recode(
            measure, "Pre-Treatment" = "bp1", "Post-Treatment" = "bp2"
            )
        )
```

Next, we create multiple density plots

```{r comment=NA, fig.width = 6, fig.height=4}
bps %>% 
    ggplot(aes(y = measure, x = bp, fill = measure)) +
    ggridges::geom_density_ridges2( col="black", alpha = .5, scale=1, 
                                    show.legend = F) +
    labs(x = "Blood pressure (mmHg)", 
         y = "Density", 
         fill = "Blood Pressure") +
    theme_bw()
```


```{r comment=NA, fig.width = 6, fig.height=4}
bps %>% 
    ggplot(aes(y = measure, x = bp, fill = measure))+
    geom_boxplot(show.legend = FALSE) +
    labs(y = NULL, 
         x = "Blood Pressure", 
         fill = "Blood Pressure") +
    coord_flip()+
    theme_light() 
```


```{r comment=NA, fig.width = 6, fig.height=4}
bps %>% 
    ggplot(aes(y = measure, x = bp, fill = measure))+
    geom_violin(show.legend = FALSE) +
    coord_flip()+
    theme_light() 
    
```

## Continuous by a single categorical variable
### Summary

We do this with one variable.
```{r comment=NA}
newdrug %>% 
    group_by(treat) %>% 
    summarize(mean.bp1 = mean(bp1),
              sd.bp1 = sd(bp1),
              var.bp1 = var(bp1),
              se.mean.bp1 = sd(bp1)/sqrt(n()),
              median.bp1 = median(bp1),
              min.bp1 = min(bp1),
              max.bp1 = max(bp1)) %>% 
    ungroup()
```


### Graph - Histogram, Boxplot, Density plot and cumulative frequency
The graphs are similar to the above so we skip them. 

## Continuous by multiple categorical variables
### Summary
This can be done as below.

```{r comment=NA}
newdrug %>% 
    group_by(treat, sex) %>% 
    summarize(mean.bp1 = mean(bp1),
              sd.bp1 = sd(bp1),
              var.bp1 = var(bp1),
              se.mean.bp1 = sd(bp1)/sqrt(n()),
              median.bp1 = median(bp1),
              min.bp1 = min(bp1),
              max.bp1 = max(bp1),
              .groups = "drop") 
```

And this can be presented in a boxplot below

```{r fig.width=6, fig.height=4}
newdrug %>% 
    ggplot(aes(y = bp1, x = sex, fill = treat)) +
    geom_boxplot()+
    labs(
        y = "Blood Pressure (mmHg)",
        x =  "Sex",
        fill = 'Treatment') +
    theme_bw()
```

