```{r echo=FALSE}
library(tidyverse)
library(magrittr)
library(gtsummary)
gtsummary::theme_gtsummary_compact()
```


# Descriptive Statistics

## Categorical Variable

For this section we use the __Titanic__ data set
 
```{r comment=NA}
titanic2 <-  
    haven::read_dta("./Data/titanic2.dta") %>% 
    mutate(sex  = haven::as_factor(sex),
           died = haven::as_factor(died),
           age  = haven::as_factor(age),
           class = haven::as_factor(class)) %>% 
    haven::zap_labels()
```

We visualize the first 6 rows of the data 

```{r comment=NA}
titanic2 %>% head()
```

And the summarize the entire as below

```{r comment=NA}
titanic2 %>% summarytools::dfSummary(graph.col = F)
```

### Single Caregorical Variable

#### Frequencies & Proportions

```{r comment=NA}
titanic2 %>% 
    gtsummary::tbl_summary(
        include = class,
        digits = class ~ c(0,1)
    ) %>% 
    gtsummary::bold_labels()
```

#### Graph - Barchart

We first summarize the data

```{r comment=NA}
bar_data <-
    titanic2 %>% 
    drop_na(class) %>% 
    count(class) %>% 
    mutate(perc = `n` / sum(`n`)) %>% 
    arrange(perc) %>%
    mutate(labels = paste(n, " (", scales::percent(perc), ")", sep=""))

bar_data
```

And the plot the barplot

```{r comment=NA, fig.width = 5, fig.height=5}
bar_data %>% 
    ggplot() +
    geom_bar(stat = "identity", 
             aes(y = n, x = class, fill = class), 
             col = "black", 
             show.legend = F) +
    geom_label(aes(y = n, label = labels, x = class), 
               vjust = 1.2,
               show.legend = FALSE, size=3.5) +
    labs(x = NULL, 
         y = "Count", 
         title = "Distribution of Class of passenger") +
    theme_bw()
```

#### Pie Chart

To do this we use the previously summarized data. Then we draw a customised Pie Chart

```{r comment=NA}
bar_data %>% 
    ggplot(aes(x = "", y = perc, fill = class)) +
    geom_col() +
    geom_label(aes(label = labels),
               position = position_stack(vjust = 0.5),
               show.legend = FALSE, size =3) +
    coord_polar(theta = "y", start=0) +
    labs(title = "Distribution of Blood Groups of study participants",
         fill = "Blood Group") +
    theme_void()
```



### Two categorical Variables

#### Frequencies & Proportions

```{r comment=NA}
titanic2 %$% 
    gmodels::CrossTable(sex, died, prop.c=F, prop.r=F, prop.t=F, prop.chisq = F)
```

#### Row percentages

```{r comment=NA}
titanic2 %$% 
    gmodels::CrossTable(sex, died, prop.c=F, prop.r=T, prop.t=F, prop.chisq = F)
```

#### Column percentages

```{r comment=NA}
titanic2 %$% 
    gmodels::CrossTable(sex, died, prop.c=T, prop.r=F, prop.t=F, prop.chisq = F)
```

#### Table Total Percentages

```{r comment=NA}
titanic2 %$% 
    gmodels::CrossTable(sex, died, prop.c=F, prop.r=F, prop.t=T, prop.chisq = F)
```

#### Bar Chart

```{r comment=NA}
titanic2 %>% 
    ggplot(aes(x = class, fill = died)) +
    geom_bar(position = position_dodge(), col = "black") +
    labs(y = "Count", x = "Class", fill = "Died",
          title = "Bar plot of outcome of passengers for each class") +
    theme_bw()
```

## Continuous variables

For this section we will use the `NewDrug_clean.dta` dataset

```{r comment=NA}
newdrug <-  
    haven::read_dta("./Data/NewDrug_clean.dta") %>% 
    mutate(sex  = haven::as_factor(sex), treat = haven::as_factor(treat)) %>% 
    haven::zap_labels() 

newdrug

newdrug %>% 
    summarytools::dfSummary(labels.col = F)
```


### Single continuous variable

#### Measures of Central tendency & Dispersion

These include mean and median and displayed below

```{r comment=NA}
newdrug %>% 
    summarise(mean.bp1 = mean(bp1), median.bp1 = median(bp1), 
              sd.bp1 = sd(bp1), min.bp1 = min(bp1), 
              max.bp1 = max(bp1), iqr = IQR(bp1)) 

```

Alternatively

```{r comment=NA}
newdrug %$% 
    psych::describe(bp1)
```

And to show the interquartile range we do the following.

```{r comment=NA}
newdrug %$% 
    psych::describe(bp1, IQR = TRUE,quant = c(.25, .75))
```

#### Graphs - Histogram 
```{r comment=NA}
newdrug %>% 
    ggplot(aes(x = bp1)) + 
    geom_histogram(bins = 7, col="black", alpha = .5, fill = "red") +
    labs(title = "Histogram of Blood Pressure before  intervention",
         x= "BP1")+
    theme_light()
```

#### Graphs - Boxplot and violin plot 

```{r comment=NA, fig.width=6, fig.height=4}
newdrug %>% 
    ggplot(aes(y = bp1)) + 
    geom_boxplot(col="black",  
                 alpha = .2, 
                 fill = "blue", 
                 outlier.fill = "black",
                 outlier.shape = 22) +
    labs(title = "Boxplot of Blood Pressure before  intervention",
         y = "BP1")+
    theme_light()
```

#### Graphs - Density plot
```{r comment=NA}
newdrug %>% 
    ggplot(aes(y = bp1)) + 
    geom_density(col="black", fill = "yellow", alpha=.6) +
    labs(title = "Density Plot of Blood Pressure before  intervention",
         y = "Blood Pressure before  intervention")+
    coord_flip() +
    theme_light()
```

#### Graphs - Cumulative Frequency plot

```{r comment=NA}
newdrug %>% 
    group_by(bp1) %>% 
    summarize(n = n()) %>% 
    ungroup() %>% 
    mutate(cum = cumsum(n)/sum(n)*100) %>% 
    ggplot(aes(y = cum, x = bp1)) +
    geom_line(col=3, size=1.2)+
    labs(title = "Cumulative Frequency Plot of Blood Pressure before  intervention",
         x = "BP1",
         y = "Cumulative Frequency")+
    theme_light() 

```

### Multiple Continuous variables

#### Measures of Central tendency & Dispersion

```{r comment=NA}
newdrug %>% 
    select(where(is.numeric)) %>% 
    psych::describe()
```

To illustrate graphing multiple continuous variables we use the 2 bp variables

```{r comment=NA}
bps <- 
    newdrug %>%
    select(bp1, bp2) %>% 
    pivot_longer(cols = c(bp1, bp2),names_to = "measure", values_to = "bp") %>% 
    mutate(measure = fct_recode(measure, 
                                "Pre-Treatment" = "bp1",
                                "Post-Treatment" = "bp2"))
```


```{r comment=NA}
bps %>% 
    ggplot(aes(y = measure, x = bp, fill = measure)) +
    ggridges::geom_density_ridges2( col="black", alpha = .5, scale=1, 
                                    show.legend = F) +
    labs(x = "Blood pressure (mmHg)", 
         y = "Density", 
         fill = "Blood Pressure") +
    theme_bw()
```


```{r comment=NA}
bps %>% 
    ggplot(aes(y = measure, x = bp, fill = measure))+
    geom_boxplot(show.legend = FALSE) +
    labs(y = NULL, 
         x = "Blood Pressure", 
         fill = "Blood Pressure") +
    coord_flip()+
    theme_light() 
```


```{r comment=NA}
bps %>% 
    ggplot(aes(y = measure, x = bp, fill = measure))+
    geom_violin(show.legend = FALSE) +
    coord_flip()+
    theme_light() 
    
```

## Continuous by single categorical variable
### Summary

We do this with one variable.
```{r comment=NA}
newdrug %>% 
    group_by(treat) %>% 
    summarize(mean.bp1 = mean(bp1),
              sd.bp1 = sd(bp1),
              var.bp1 = var(bp1),
              se.mean.bp1 = sd(bp1)/sqrt(n()),
              median.bp1 = median(bp1),
              min.bp1 = min(bp1),
              max.bp1 = max(bp1)) %>% 
    ungroup()
```

Alternatively we can use
```{r comment=NA}
newdrug %>% 
    select(where(is.numeric), treat) %>% 
    psych::describeBy(group = "treat")
```

### Graph - Histogram, Boxplot, Density plot and cumlative frequency
The graphs are similar to the aboveso we skip them. 

## Continuous by multiple categorical variables
### Summary
This csn be done as below.

```{r comment=NA}
newdrug %>% 
    group_by(treat, sex) %>% 
    summarize(mean.bp1 = mean(bp1),
              sd.bp1 = sd(bp1),
              var.bp1 = var(bp1),
              se.mean.bp1 = sd(bp1)/sqrt(n()),
              median.bp1 = median(bp1),
              min.bp1 = min(bp1),
              max.bp1 = max(bp1)) %>% 
    ungroup()
```

