# Importing data into R

```{r echo = F, include=F, meassage = F, warning=F}
library(huxtable)
library(tidyverse)
library(magrittr)
```

In this chapter, we discuss ways of getting data into R, either by directly entering it into R or importing it from another software. In R, a data frame is the data structure desirable for data analysis. With the advent of the `tidy data`, a tibble is not the predominant data structure being used. For this section, we will read the data below but from different formats.

```{r echo =FALSE}
tibble(
    name = c("Ama", "Yakubu", "John"), 
    sex = c("Female", "Male", "Male"),
    age = c(12, 9, 4),
    school = c("JHS", "Primary", "Creche")
    )
```

## Direct entry into 
We first use the `data.frame` function from the base package to create the data frame.

```{r}
data.frame(
    name = c("Ama", "Yakubu", "John"), 
    sex = c("Female", "Male", "Male"),
    age = c(12, 9, 4),
    school = c("JHS", "Primary", "Creche")
    )
```


Below we first describe how to manually enter data into R. We aim to create a tibble by using the `tibble` function.

```{r}
tibble(
    name = c("Ama", "Yakubu", "John"), 
    sex = c("Female", "Male", "Male"),
    age = c(12, 9, 4),
    school = c("JHS", "Primary", "Creche")
    )
```


## Text files

## Microsoft Excel

## SPSS files

## dbf files

## Other files