# Importing data into R

```{r echo = F, include=F, meassage = F, warning=F}
library(huxtable)
library(tidyverse)
library(magrittr)
```

In this chapter, we discuss ways of getting data into R, either by directly entering it into R or importing it from another software. In R, a data frame is the data structure desirable for data analysis. With the advent of the `tidy data`, a tibble is now the predominant data structure being used. For this section, we will be reading in various file formats and presenting them as a tibble.

## Using data in R packages
Many packages in R come with data that can be used for practice. To be able to use a dataset in a specific package, that package first has to be installed. For instance, to be able to use the data `Oswego` native to the package `epiDisplay`, we first ensure the package is installed by running the command line below:


```{r echo = T, eval=F}
install.packages("epiDisplay")
```

Note that this will install the package and any other packages that the `epiDisplay` package depends on.

The next step will be to make the data available in the R session as below

```{r echo=T}
data("Oswego", package = "epiDisplay")
```

Now that the data is available in the working environment, we can visualise the first 6 rows below

```{r}
Oswego %>% head()
```

And finally visualise the details of the data by using the `help()` function as below

```{r eval = F}
help(Oswego)
```


## Direct entry into R
We first use the `data.frame()` function from the base package to create the data frame.

```{r}
data.frame(
    name = c("Ama", "Yakubu", "John"), 
    sex = c("Female", "Male", "Male"),
    age = c(12, 9, 4),
    school = c("JHS", "Primary", "Creche")
    )
```


Below we first describe how to manually enter data into R. We aim to create a tibble by using the `tibble` function.

```{r}
tibble(
    name = c("Ama", "Yakubu", "John"), 
    sex = c("Female", "Male", "Male"),
    age = c(12, 9, 4),
    school = c("JHS", "Primary", "Creche")
    )
```

## R data file
When working in R, a frequent mode of storage of data is as an .Rdata file. This preserves the structure and environment of the data. Below we will read an already saved .Rdata file.

```{r}
load(file = "./Data/data1.Rdata")
ls()
```
We then visualise the first 4 rows of the single data within the loaded file called `data1_stata`

```{r}
data1_stata %>% head(n=4)
```


## Text files
The first file format that we are going to read from is a flat file or text file. These usually have the extension .txt. The data in these files could be separated by various delimiters. These include tabs, commas, spaces, etc. In this section, we will read in one with a tab delimiter as a prototype as the rest will be similar.

```{r}
read_delim(
    file = "./Data/bpA.txt", 
    delim = "\t", 
    col_types = c("c", "c", "d", "d")
    )
```

The last file to be read in this subsection is a comma-delimited text file

```{r}
read_delim(
    file = "./Data/blood.txt", 
    delim = ",", 
    col_types = c("c", "d", "d", "d", "d", "d", "d")
    ) %>% 
    head(n=4)
```

Comma-delimited files with extension .csv can also be imported with the commnands

```{r}
read_csv(
    file = "./Data/blood.txt",
    col_types = c("c", "d", "d", "d", "d", "d", "d")) %>% 
    head(n=4)
```

## Microsoft Excel
Probably the most common format for transferring data is Microsoft Excel. There are two versions of Excel with extensions .xls and .xlsx. Below reading in the .xlsx is demonstrated using the `readxl` package.

```{r}
readxl::read_xlsx(path = "./Data/data1.xlsx") %>% 
    head(n=4)
```

## SPSS files
Files from SPSS are usually saved with the extension of .sav. Below we read an SSPSS data file using the `haven` package

```{r}
haven::read_sav(file = "./Data/data1.sav") %>% 
    head(n=4)
```

## Stata files
Stata files, similar to SPSS data files can be imported using the `haven` package. This is illustrated below

```{r}
haven::read_dta(file = "./Data/data1.dta") %>% 
    head(n=4)
```

## SAS files
The `haven` package also offers the ability to read into R a SAS data file. This is illustrated below

```{r}
haven::read_sas(data_file = "./Data/data1.sas7bdat") %>% 
    head(n=4)
```

## Conclusion

In this section, we have learned how to import data into R from various file formats and programs. In the next section, we will learn about how to export data in R to other file formats. 