# Ordinal logistic regression

```{r include=FALSE}
library(tidyverse)
```

## Importing data

In this presentation we analyse a dataset using ordinal logistic regression. 
We begin by reading in the data and selecting our desired subset.

```{r comment = NA}
dataF <- 
    dget("./Data/anemia_data") %>% 
    select(sid, anemia_cat, community, fever, sex,
           famsize, moccup2, foccup2, hosp_visit)
```

We then view a summary of the data

```{r comment = NA}
dataF %>% glimpse()

```

Note that the `anemia_cat` variable is an ordered factor variable. 
For completeness the single missing observation for the variable `hosp_adm` 
will be recoded to `No`.

```{r comment = NA}
dataF <- dataF %>% 
    mutate(hosp_visit = forcats::fct_explicit_na(hosp_visit, na_level = "No"))
summary(dataF)
```

## Model specification

Now we begin the ordinal regression by fixing the first model, the Null model.

```{r comment = NA}
Model_0 <- ordinal::clm(anemia_cat ~ 1, data = dataF, link = "logit")
summary(Model_0)
```

Subsequently we introduce the fever variable as independent and express the 
results as OR with 95%CI

```{r comment=NA}
Model_1 <- ordinal::clm(anemia_cat ~ fever, data = dataF, link = "logit")

broom::tidy(Model_1, conf.int = TRUE, exponentiate = TRUE)%>% 
    flextable::as_flextable() %>% 
    flextable::colformat_double(
        j = c("estimate", "std.error", "statistic", "p.value", 
              "conf.low", "conf.high"), 
        digits = 3)
```

Results indictes a significant association between fever and the degree of anemaia (OR=1.46, 95%CI: 1.02 to 2.09). Performing an anova test to see if there exist a difference between the 2 models yields.

```{r comment = NA}
anova(Model_0, Model_1)
```

The results indicates adding fever to the Null model significantly improves the null model.

Next we add the community variable

```{r comment = NA}
Model_2 <- 
     ordinal::clm(anemia_cat ~ fever + community, data = dataF, link = "logit")

broom::tidy(Model_2, conf.int = TRUE, exponentiate = TRUE)%>% 
    flextable::as_flextable() %>% 
    flextable::colformat_double(
        j = c("estimate", "std.error", "statistic", "p.value", 
              "conf.low", "conf.high"), 
        digits = 3)
```

## Checking proportioinal odds assumption for the model

Here we check the proportional odd assumption for out second model

```{r comment = NA}
ordinal::nominal_test(Model_2)
```

The significant p-value for the `community` variable indicates a breech of the proportional odd assumption

## Prediction
In this section we will use the model created above to predict an observation being in a specific anemia severity group. First we bigin by forming the prediction data we call `newData`.

```{r comment = NA}
NewData <- expand.grid(community = levels(dataF$community),
                       fever = levels(dataF$fever))
NewData
```

We now predict the probability that the specific predictor combination falls within the specific outcome category (anemia category)

```{r comment=NA}
(preds <- predict(Model_2, newdata = NewData, type = "prob"))
```

For better visualiszation we bind the original data with the predictions

```{r comment=NA}
bind_cols(NewData, preds$fit) %>% 
    kableExtra::kbl(caption = "Probabilities", booktabs = T, digits = 3) %>%
    kableExtra::kable_classic(full_width = F, html_font = "Cambria") %>% 
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualising the model

Below we visualize the model by using the `MASS` and `effects` packages. We begin by fitting the model again with `polr` function.

```{r comment = NA}
pol_model.1 <- MASS::polr(anemia_cat ~ community, data = dataF)
pol_model.2 <- MASS::polr(anemia_cat ~ fever*community, data = dataF)
```

And the we visualise the probablitly of having the various forms of anemia giving one belongs to the various groups.


```{r  comment = NA}
M1 <- effects::Effect(focal.predictors = "community", mod=pol_model.1)
M2 <- effects::Effect(focal.predictors = c("community", "fever"), mod=pol_model.2)
plot(M1)
plot(M2)
```